---
title: "Data cleaning, Categorization, Favorite category filter"
author: "Ugnius, Bart, Kalle, Eline"
date: "2025-01-31"
output: html_document
---
## Loading packages and data

Loading Packages
```{r}
install.packages(c("tidyverse", "tidytext", "caret", "text2vec", "dplyr", "stringr"))
```

```{r}
library(tidyverse)
library(tidytext)
library(caret)
library(text2vec)
library(dplyr)
library(stringr)
library(tidyr)
library(recommenderlab)
```

Load the data
```{r}
rm(list = ls())

Online_Retail_Data = read.csv("Online_Retail_Data.csv")
colnames(Online_Retail_Data)[colnames(Online_Retail_Data) == "ļ.æInvoiceNo"] <- "InvoiceNumber"
```


## Data cleaning

Clean the data
```{r}
data <- Online_Retail_Data %>%
  filter(
    grepl("^[A-Z ]+$", Description) &  
     !is.na(CustomerID) &
    UnitPrice >= 0.01  & UnitPrice <= 15,
    StockCode != "POST", 
    Country == "United Kingdom"
  )

# Apply the transformations
data <- data %>%
  # Remove Customers with Fewer than 15 Transactions
  group_by(CustomerID) %>%
  filter(n() >= 15) %>%
  ungroup() %>%
  # Remove Observations with Quantity > 250
  filter(Quantity <= 250) %>%
  # Remove Items Purchased Fewer than 100 Times
  group_by(Description) %>%
  filter(sum(Quantity) >= 100) %>%
  ungroup()

data <- data %>%
  mutate(
    InvoiceDate = as.POSIXct(InvoiceDate, format = "%m/%d/%y %H:%M"), # Parse the date and time
    Date = as.Date(InvoiceDate),                                     # Extract the date part
    Time = format(InvoiceDate, "%H:%M:%S"),                          # Extract the time part
    Month = format(Date, "%B")                                       # Extract the month name
  ) %>%
  select(-InvoiceDate) %>%                                           # Remove the original 
  # Ensure Month is ordered from January to December
  mutate(Month = factor(Month, levels = month.name))

# View the cleaned dataset
print(head(data))
```

Removing all the  transactions that are returned and earlier bought items that are returned. 
```{r}
# Step 1: Split the dataset into purchases and returns
purchases <- data %>%
  filter(Quantity > 0)

returns <- data %>%
  filter(Quantity < 0)

# Step 2: Identify matched transactions (same ConsumerID, Description, and absolute Quantity)
matched_transactions <- purchases %>%
  inner_join(
    returns,
    by = c("CustomerID", "Description"),
    suffix = c("_purchase", "_return")
  ) %>%
  filter(abs(Quantity_purchase) == abs(Quantity_return))  # Match on absolute quantity

# Step 3: Remove both purchase and return transactions from the dataset
data <- data %>%
  anti_join(matched_transactions %>%
              select(CustomerID, Description, Quantity_purchase, Quantity_return) %>%
              pivot_longer(cols = starts_with("Quantity"),
                           names_to = NULL,
                           values_to = "Quantity"),
            by = c("CustomerID", "Description", "Quantity"))


# step 4: remove all the transactions that are earlier bought and also returned. 
data <- data %>% filter(Quantity > 0)
```



## Categorization

Insight in the different Descriptions
```{r}
description_summary <- data %>%
  group_by(Description) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

print(description_summary)
```


Rules-based dictionary categorization 
```{r}
categories <- list(
  # Home Decor
  "Lighting and Ambiance" = c(
    "LANTERN" = 1, "T-LIGHT" = 1, "CANDLE" = 1, "CANDLESTICK" = 1, 
    "CANDLEHOLDER" = 1, "TERRARIUM" = 1, "HOME" = 1, "DOORSTOP" = 1, "DOORMAT" = 1, "COATHANGER" = 1, "NAMASTE" = 1, "INCENSE" = 2
  ),
  
  "Wall and Hanging Decor" = c(
    "FRAME" = 1, "PICTURE" = 1, "MIRROR" = 1, "WALL" = 2, "GARLAND" = 2, "MAGNET" = 1, "MAGNETS" = 1,  "HOOK" = 1, "PHOTO" = 1),
  
  "Soft Furnishings" = c(
    "RUG" = 1, "CURTAIN" = 1, "PILLOW" = 1, "CUSHION" = 1, "BLANKET" = 2
  ),
  
  "Decorative Accents" = c(
    "SCULPTURE" = 2, "ART" = 2, "VASE" = 2, "BASKET" = 3, "STATUE" = 2, 
    "BLOCKWORD" = 1, "WOODLETTERS" = 1, "BIRDORNAMENT"= 1, "SIGN" = 1
  ),
  
  "Sculptures and Statues" = c(
    "STATUE" = 1, "SCULPTURE" = 1, "FIGURINE" = 2 
  ),
  
  "Floral and Nature-Inspired Decor" = c(
    "VASE" = 1, "PLANT" = 1, "FLOWER" = 1, "TERRARIUM" = 1
  ),
  
  "Seating and Lounging" = c(
    "CHAIR" = 1, "SOFA" = 1, "BENCH" = 2, "ARMCHAIR" = 2, "OTTOMAN" = 2
  ),
  
  "Furniture and Tables" = c(
    "TABLE" = 1, "DESK" = 1, "COFFEE" = 1, "SIDE" = 2, "BOOKCASE" = 2, 
    "DRESSER" = 2
  ),
  
  "Storage Solutions" = c(
    "CABINET" = 1, "SHELF" = 1, "WARDROBE" = 2, "DRAWER" = 2, "STORAGE" = 3
  ),
  
  "Containers and Boxes" = c(
    "BOX" = 1, "BASKET" = 1, "BIN" = 2, "CASE" = 2, "TRUNK" = 3, "JAR" = 1, "JARS" = 1
  ),
  
  "Lighting" = c(
    "LIGHT" = 1, "LAMP" = 1, "LED" = 1, "BULB" = 1, "CHANDELIER" = 2, "PENDANT" = 3
  ),
  
  "Cleaning Supplies" = c(
    "CLOTH" = 1, "SPONGE" = 1, "MOP" = 1, "BRUSH" = 1, "DUSTER" = 1, 
    "CLEANER" = 1, "SPRAY" = 1, "SOAP" = 2
  ),

  # Office Supplies
  "Writing and Paper Supplies" = c(
    "PEN" = 1, "NOTEBOOK" = 1, "PAPER" = 1, "MARKER" = 2, 
    "RULER" = 3, "ERASER" = 3, "PENCIL" = 1, "HIGHLIGHTER" = 2, 
    "STICKY" = 3, "JOURNAL" = 2, "LABEL" = 3, "CARD" = 3, "MEMOBOARD" = 1
  ),
  
  "Organizational and Desk Tools" = c(
    "CLIP" = 1, "STAPLER" = 2, "BINDER" = 2, "FOLDER" = 3, 
    "CALENDAR" = 3, "TAPE" = 3, "PAPERCLIP" = 3, "RUBBER" = 4, 
    "PUNCH" = 4
  ),

  # Clothing
  "Clothing" = c(
    "T-SHIRT" = 1, "DRESS" = 1, "COAT" = 1, "JACKET" = 1, "SHIRT" = 2,
    "PANTS" = 2, "SOCKS" = 2, "SHOES" = 1, "BOOTS" = 2, "SKIRT" = 2,
    "VEST" = 3, "SHORTS" = 3, "HOODIE" = 3, "BLOUSE" = 3,
    "JUMPSUIT" = 3, "GOWN" = 2, "SUIT" = 2, "PAJAMAS" = 4
  ),
  
  "Fashion Accessories" = c(
    "SCARF" = 1, "GLOVES" = 1, "BELT" = 2, "HAT" = 2, 
    "SUNGLASSES" = 2, "HEADBAND" = 3, "EYEWEAR" = 3
  ),
  
  "Functional Items" = c(
    "BAG" = 1, "WALLET" = 2, "UMBRELLA" = 2, "KEYRING" = 2, 
    "PURSE" = 1, "MONEYCLIP" = 2, "CANE" = 3, "HANGER" = 3, 
    "SEWINGKIT" = 1, "JUMBOSHOPPER" = 3
  ),

  # Jewellery
  "Jewellery" = c(
    "NECKLACE" = 1, "EARRINGS" = 1, "BRACELET" = 1, "WATCH" = 1, 
    "RING" = 1, "PENDANT" = 2, "BROOCH" = 3, "ANKLET" = 2, 
    "TIARA" = 2, "BANGLE" = 2, "PEARL" = 1, "DIAMOND" = 1, 
    "JEWELLERY" = 1, "JEWEL" = 1
  ),

  # Creative
  "Art Supplies" = c(
    "PAINT" = 1, "BRUSH" = 1, "CRAYON" = 2, "MARKER" = 2, 
    "PENCIL" = 3, "ART" = 2, "SKETCH" = 3, "PASTEL" = 4, 
    "STAMP" = 4, "STENCIL" = 3, "ORIGAMI" = 1
  ),
  "Crafting Materials" = c(
    "GLUE" = 1, "SCISSORS" = 1, "YARN" = 1, "NEEDLE" = 2, "FABRIC" = 2, 
    "THREAD" = 3, "BUTTON" = 3, "KNITTED" = 2, "CLAY" = 1, "SCISSOR" = 1,
    "CHALKBOARD" = 1
  ),
  
  "Craft Decorative Accessories" = c(
    "EMBROIDERED" = 1, "PATCH" = 3, "FLOWER" = 1, "RIBBONS" = 2, 
    "LACE" = 2, "BEADS" = 2, "STICKERS" = 1, "STICKER" = 1 
  ),

  # Party
  "Party Decorations" = c(
    "BALLOON" = 1, "CONFETTI" = 1, "BANNER" = 1, 
    "STREAMERS" = 2, "DECORATION" = 2, "TIARAS" = 1, "BIRTHDAY" = 1, 
    "CONGRATULATIONS" = 1, "SOMBRERO" = 1
  ),
  
  "Party Food and Ware" = c(
    "CANDY" = 1, "CAKE" = 1, "CUPCAKE" = 2, "CUP" = 4
  ),
  
  "Gifts" = c(
    "GIFT" = 1, "PRESENT" = 1, "WRAP" = 2, "BOW" = 4
  ),

  # Christmas
  "Christmas" = c(
    "CHRISTMAS" = 1, "STAR" = 2, "ANGEL" = 1, 
    "GINGERBREAD" = 1, "FAIRY" = 2
  ),
  
  # Easter
  "Easter" = c(
    "EASTER" = 1, "BUNNY" = 2
  ),

  # Children
"Children's Items & Toys" = c( 
  "TOYS" = 1, "TOY" = 1, "BLOCK" = 1, "DOLL" = 1, "DINOSAUR" = 1, "TEDDY" = 1, 
  "BABY" = 1, "PRINCESS" = 1, "MAGIC" = 2, "SKULLS" = 2, "SKULL" = 2, "CIRCUS" = 2, 
  "BUNNIES" = 1, "SPACEBOY" = 1, "JIGSAW" = 2, "KIDS" = 1, "WITCH" = 2, 
  "PINK" = 1, "SKIPPING" = 2, "PIRATE" = 2, "HELICOPTER" = 1, "SHARK" = 2, 
  "GIRL" = 2, "FARM" = 2, "PUPPET" = 2, "SPACE" = 1, "DRAGONFLY" = 2, 
  "BUTTERFLY" = 2, "DRAGONS" = 2, "SHEEP" = 2, "RABBIT" = 1, "CHILDREN" = 1, 
  "CHILDRENS" = 1, "DOLPHIN" = 2, "DOLLY" = 1
  ),

  # Outdoor
"Garden" = c(
  "PLANTER" = 1, "POT" = 1, "SHOVEL" = 1, "RAKE" = 2, "WATERING" = 2,
  "GARDEN" = 1, "SEEDS" = 2, "SOIL" = 3, "COMPOST" = 3, "TRELLIS" = 3,
  "HOSE" = 3, "SPADE" = 4, "MULCH" = 4, "SHREDDER" = 4, "FOOT" = 4,
  "BLOOM" = 2, "HERBS" = 3, "GROW" = 2, "CULTIVATOR" = 3, "PARASOL" = 1, 
  "BIRDHOUSE" = 1, "GLOVE" = 4, "HAMMOCK" = 1, "FARMYARD" = 1, "HEN" = 2
  ),

"Outdoor Gear" = c(
  "BICYCLE" = 1, "FISHING" = 2, "CAMPING" = 2, "SLEEPING" = 2, 
  "STOVE" = 2, "CHAIR" = 3, "RUG" = 3, "JACKET" = 2, "FIRE" = 2, 
  "BINOCULARS" = 3, "GRILL" = 4, "PICNIC" = 4, "COOLER" = 4, 
  "FUEL" = 3, "SEAT" = 4, "PONCHO" = 1, "RUCKSACK" = 1, "HANDWARMER" = 1
  ),

# Travel
"Travel" = c(
  "LUGGAGE" = 1, "PASSPORT" = 1, "TENT" = 1, "BACKPACK" = 1, 
  "CAMPING" = 1, "CAMP" = 1, "TRAVEL" = 1
  ),

"Electronics & Gadgets" = c(
  "CHARGER" = 1, "HEADPHONES" = 1, "PHONE" = 1, "LAPTOP" = 1, 
  "TABLET" = 2, "CAMERA" = 2, "LIGHT" = 2, "LAMP" = 2, 
  "KEYBOARD" = 2, "RADIO" = 3, "CLOCK" = 3, "SCREEN" = 3, 
  "MICROPHONE" = 3, "BATTERY" = 3, "REMOTE" = 3, "MOUSE" = 4, 
  "CAMCORDER" = 4, "FLASH" = 4, "DRIVE" = 4, "CALCULATOR" = 4, 
  "USB" = 4, "DEVICE" = 4, "WIRE" = 4, "ROBOT" = 4, 
  "GADGET" = 4, "GIZMO" = 4, "TIMER" = 4, "GEAR" = 4, "CABLE" = 4,
  "MOBILE" = 1, "FAN" = 1
  ),

# Bathroom
"Beauty & Personal Care" = c(
  "SOAP" = 1, "LOTION" = 1, "SHAMPOO" = 1, "PERFUME" = 1, 
  "MAKEUP" = 1, "BRUSH" = 2, "CREAM" = 2, "COSMETICS" = 2, 
  "HAIR" = 2, "MASK" = 2, "MIRROR" = 3, "BATH" = 3, 
  "LIP" = 3, "NAIL" = 3, "FACIAL" = 3, "EXFOLIANT" = 3, 
  "SCRUB" = 3, "TONER" = 4, "HAND" = 4, "MINT" = 4, "POMADE" = 4, "SHOWERCAP" = 1
  ),

"Medical Supplies" = c(
  "MASK" = 1, "SANITIZER" = 1, "GLOVES" = 1, "FIRST AID" = 1, 
  "MEDICAL" = 1, "PLASTERS" = 2, "BANDAGE" = 2, "ANTISEPTIC" = 2, 
  "SURGICAL" = 2, "ALCOHOL" = 3, "DISINFECTANT" = 3, "STERILE" = 3, 
  "PILLS" = 3, "MEDICINE" = 3, "SYRINGE" = 4, "ASPIRIN" = 4, 
  "TABLET" = 4, "CURE" = 4, "HEALTH" = 4, "CARE" = 4
  ),

# Bedroom
"Bedroom" = c(
  "ALARMCLOCK" = 1, "BED" = 1
  ),

# Kitchen Clothing
"Kitchen Clothing" = c(
  "APRON" = 1, "CHEF" = 2, "KITCHEN" = 2, "KITCHENCLOTH" = 2, 
  "KITCHENGLOVE" = 2, "COOKINGHAT" = 3, "TEATOWEL" = 3, "POTHOLDER" = 4
  ),

"Kitchenware" = c(
  "MUG" = 1, "EGG" = 1, "TEA" = 1, "PLATE" = 1, "BISCUIT" = 2, 
  "BISCUITTIN" = 2, "SQUEEZER" = 2, "FOODCOVER" = 2, 
  "DINNERTRAY" = 2, "MATCHES" = 3, "ORGANISER" = 3, "GLASS" = 1,
  "JAMJAR" = 3, "POPCORN" = 2, "FRIDGE" = 2, "DINNER" = 2, 
  "BREAKFAST" = 1, "FLASK" = 3, "TRAY" = 3, "WATERBOTTLE" = 1, "LUNCH" = 1, 
  "JUG" = 3, "JUGS" = 3, "BOTTLE" = 2, "BOTTLES" = 2, "JAM" = 2, "PAN" = 2, "CHOPPING" =2 ))

seasonal_keywords <- list(
  "Christmas" = c("CHRISTMAS", "WINTERHOLIDAY", "NEWYEAR", "SANTA", "STOCKING", "REINDEER", "TREE", "ORNAMENTS", "SNOWFLAKE", "MISTLETOE"),
  "Easter" = c("EASTER", "BUNNY", "EGG", "BLOSSOM", "CROSS", "RESURRECTION"),
  "Halloween" = c("HALLOWEEN", "PUMPKIN", "SPOOKY", "GHOST", "WITCH", "SKULL", "SPIDERWEB", "COSTUME", "BAT"),
  "Valentine's Day" = c("VALENTINE", "HEART", "CUPID", "LOVE", "ROMANCE", "ROSE", "CHOCOLATES"),
  "Thanksgiving" = c("THANKSGIVING", "TURKEY", "HARVEST", "FEAST", "AUTUMNLEAVES", "CORNUCOPIA", "STUFFING", "CRANBERRY")
)


categorize <- function(description, categories) {
  matched_categories <- lapply(names(categories), function(cat) {
    # Use \\b to ensure exact matches for words (e.g., "WALL" won't match "WALLET")
    matched_keywords <- names(categories[[cat]])[str_detect(description, paste0("\\b", names(categories[[cat]]), "\\b"))]
    if (length(matched_keywords) > 0) {
      priority <- min(unlist(categories[[cat]][matched_keywords]))
      return(list(category = cat, priority = priority))
    } else {
      return(NULL)
    }
  })

  matched_categories <- Filter(Negate(is.null), matched_categories)

  if (length(matched_categories) > 0) {
    best_match <- matched_categories[[which.min(sapply(matched_categories, `[[`, "priority"))]]
    return(best_match$category)
  }

  return("Other")
}

identify_seasonality <- function(description, seasonal_keywords) {
  for (season in names(seasonal_keywords)) {
    if (any(str_detect(description, seasonal_keywords[[season]]))) {
      return(season)
    }
  }
  return("Non-Seasonal")
}

data$Description <- toupper(data$Description)

# Apply categorization function to data
data$Category <- sapply(data$Description, categorize, categories)
data$Seasonality <- sapply(data$Description, identify_seasonality, seasonal_keywords)
```

Insight in unlabeled ('Other') Descriptions
```{r}
other_category_data <- data[data$Category == "Other", ]
print(other_category_data)

# Categories
table(data$Category)
```
Plot for the Categories
```{r}
category_counts <- data %>%
  count(Category, name = "Count")

ggplot(category_counts, aes(x = reorder(Category, Count), y = Count, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Category Distribution",
    x = "Category",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8, hjust = 1), # Slightly reduced text size for readability
    axis.title.x = element_text(size = 10),          # Adjusted x-axis title size
    axis.title.y = element_text(size = 10, angle = 0, vjust = 0.5), # Center the y-axis label
    plot.margin = unit(c(0.5, 1, 0.5, 4), "cm"),     # Adjust margins
    legend.position = "none"                         # Remove legend
  ) +
  coord_flip() +
  scale_y_continuous(breaks = seq(0, 40000, by = 5000)) # Adjust y-axis breaks

# Save the plot with a larger width
ggsave("category_distribution_wide.png", width = 12, height = 8, dpi = 300)
```

Visualization of the amount of Items sold per Category
```{r}
summary(data$Quantity)
head(data$Quantity)

# Aggregate total quantity sold per category
category_sales <- data %>%
  group_by(Category) %>%
  summarise(TotalQuantity = sum(Quantity)/ 10, .groups = "drop") %>%
  arrange(desc(TotalQuantity))

category_sales

# Create the bar plot
bar_plot <- ggplot(category_sales, aes(x = reorder(Category, TotalQuantity), y = TotalQuantity, fill = Category)) +
  geom_bar(stat = "identity") +
  coord_flip() +  # Flip bars for better readability
  scale_y_continuous(labels = scales::comma_format(scale = 1)) + 
  labs(
    title = "Total Items Sold Per Category",
    x = "Category",
    y = "Total Quantity Sold"
  ) +
  theme_minimal() +
  theme(legend.position = "none")  # Remove legend since color is just for categories

bar_plot

# Save the plot as a PDF in your Documents folder
ggsave(
  filename = "~/Documents/total_items_sold_per_category.pdf",  # File path and name
  plot = bar_plot,  # The plot object to save
  width = 8,  # Width in inches
  height = 6,  # Height in inches
  dpi = 300  # Resolution (not very relevant for PDFs but ensures clarity)
)
```

F1 Score for Validation Categorization
```{r}
calculate_f1_score <- function(actual, predicted) {
  true_positive <- sum(actual == predicted & actual != "Other")
  false_positive <- sum(actual != predicted & predicted != "Other")
  false_negative <- sum(actual != predicted & actual != "Other")

  precision <- true_positive / (true_positive + false_positive)
  recall <- true_positive / (true_positive + false_negative)

  if (is.nan(precision) | is.nan(recall)) {
    return(0)
  }

  f1_score <- 2 * ((precision * recall) / (precision + recall))
  return(f1_score)
}

# Simulate splitting data into folds for cross-validation
data_split <- split(data, sample(1:4, nrow(data), replace = TRUE))

f1_scores <- sapply(data_split, function(split_data) {
  actual <- data$Category[match(split_data$Description, data$Description)]
  predicted <- split_data$Category
  calculate_f1_score(actual, predicted)
})

print(f1_scores)
```



## Benchmark Model: Generating Random Item Recommendations for a Customer
```{r}
generate_random_recommendations <- function(customer_id, train_data, N = 5) {
  # Get all unique items from the training data
  all_items <- unique(train_data$Description)
  
  # Select N random items
  random_items <- sample(all_items, N, replace = FALSE)
  
  return(random_items)}
```

Hit Ratio Calculation
```{r}
calculate_hit_ratio_random <- function(test_data, train_data, N = 5) {
  hits <- 0
  total_users <- length(unique(test_data$CustomerID))
  
  for (customer_id in unique(test_data$CustomerID)) {
    # Actual items purchased in the test set
    actual_items <- unique(test_data$Description[test_data$CustomerID == customer_id])
    
    # Randomly recommended items
    recommended_items <- generate_random_recommendations(customer_id, train_data, N)
    
    # Check for a hit (at least one match)
    if (length(intersect(recommended_items, actual_items)) > 0) {
      hits <- hits + 1
    }
  }
  
  # Compute Hit Ratio
  hit_ratio <- hits / total_users
  return(hit_ratio)
}

# Compute Hit Ratio for the random recommender
hit_ratio_random <- calculate_hit_ratio_random(test_data, train_data, N = 5)
cat("Hit Ratio (Random Recommendations):", hit_ratio_random, "\n")
```

Precision and Recall Calculation
```{r}
calculate_precision_recall_random <- function(test_data, train_data, N = 5) {
  total_precision <- 0
  total_recall <- 0
  total_users <- length(unique(test_data$CustomerID))
  
  for (customer_id in unique(test_data$CustomerID)) {
    # Actual purchased items
    actual_items <- unique(test_data$Description[test_data$CustomerID == customer_id])
    
    # Random recommendations
    recommended_items <- generate_random_recommendations(customer_id, train_data, N)
    
    # Compute True Positives (TP)
    true_positives <- length(intersect(recommended_items, actual_items))
    
    # Compute Precision
    precision <- ifelse(length(recommended_items) > 0, true_positives / length(recommended_items), 0)
    
    # Compute Recall
    recall <- ifelse(length(actual_items) > 0, true_positives / length(actual_items), 0)
    
    # Accumulate metrics
    total_precision <- total_precision + precision
    total_recall <- total_recall + recall
  }
  
  # Compute average Precision and Recall
  avg_precision <- total_precision / total_users
  avg_recall <- total_recall / total_users
  
  return(list(precision = avg_precision, recall = avg_recall))
}

# Compute Precision and Recall for the random recommender
precision_recall_random <- calculate_precision_recall_random(test_data, train_data, N = 5)
cat("Precision (Random Recommendations):", precision_recall_random$precision, "\n")
cat("Recall (Random Recommendations):", precision_recall_random$recall, "\n")
```

NDCG Calculation
```{r}
calculate_ndcg_random <- function(test_data, train_data, N = 5) {
  total_ndcg <- 0
  total_users <- length(unique(test_data$CustomerID))
  
  for (customer_id in unique(test_data$CustomerID)) {
    # Actual purchased items
    actual_items <- unique(test_data$Description[test_data$CustomerID == customer_id])
    
    # Random recommendations
    recommended_items <- generate_random_recommendations(customer_id, train_data, N)
    
    # Skip customers with no recommendations or no actual items
    if (length(recommended_items) == 0 || length(actual_items) == 0) {
      next
    }
    
    # Compute DCG
    dcg <- sum(sapply(1:length(recommended_items), function(rank) {
      if (recommended_items[rank] %in% actual_items) {
        return(1 / log2(rank + 1))
      } else {
        return(0)
      }
    }))
    
    # Compute IDCG (Ideal DCG)
    ideal_dcg <- sum(sapply(1:min(length(actual_items), N), function(rank) {
      return(1 / log2(rank + 1))
    }))
    
    # Compute NDCG
    ndcg <- ifelse(ideal_dcg > 0, dcg / ideal_dcg, 0)
    
    # Accumulate NDCG
    total_ndcg <- total_ndcg + ndcg
  }
  
  # Compute average NDCG
  avg_ndcg <- total_ndcg / total_users
  return(avg_ndcg)
}

# Compute NDCG for the random recommender
ndcg_random <- calculate_ndcg_random(test_data, train_data, N = 5)
cat("NDCG (Random Recommendations):", ndcg_random, "\n")

```


## Favorite Category Recommender System

Data per ConsumerID
```{r}
# Aggregate data by ConsumerID
consumer_summary <- data %>%
  group_by(CustomerID) %>%
  
  summarise(
    total_quantity = sum(Quantity, na.rm = TRUE),           # Total quantity purchased
    unique_items = n_distinct(Description),                 # Number of unique items bought
    total_spent = sum(UnitPrice * Quantity, na.rm = TRUE),  # Total spending
    total_transactions = n()                                # Number of transactions
  )

# View the aggregated data
print(head(consumer_summary))
```

Favorite category per Customer
```{r}
# Group by CustomerID and Category, and calculate total quantity for each category
customer_category_summary <- data %>%
  group_by(CustomerID, Category) %>%
  summarise(total_quantity = sum(Quantity, na.rm = TRUE), .groups = "drop")

# Find the favorite category for each customer
favorite_category <- customer_category_summary %>%
  group_by(CustomerID) %>%
  slice_max(total_quantity, n = 1, with_ties = FALSE) %>%  # Select the category with the highest quantity
  ungroup()

# View the results
print(head(favorite_category))
```

Extend favorite category
```{r}
# Group by CustomerID and Category, and calculate total quantity for each category
customer_category_summary <- data %>%
  group_by(CustomerID, Category) %>%
  summarise(total_quantity = sum(Quantity, na.rm = TRUE), .groups = "drop")

# Find the favorite category for each customer
favorite_category <- customer_category_summary %>%
  group_by(CustomerID) %>%
  slice_max(total_quantity, n = 1, with_ties = FALSE) %>%  # Select the category with the highest quantity
  ungroup() %>%
  rename(`Category_quantity` = total_quantity)  # Rename total_quantity to category_quantity

# Add the total quantity of items bought by each CustomerID
favorite_category <- favorite_category %>%
  left_join(
    data %>%
      group_by(CustomerID) %>%
      summarise(Total_quantity = sum(Quantity, na.rm = TRUE), .groups = "drop"),
    by = "CustomerID"
  )

# View the results
print(head(favorite_category))
```

Merge two datasets
```{r}
# Merge the datasets, adding only Category and Category_quantity to consumer_summary
combined_summary <- consumer_summary %>%
  left_join(
    favorite_category %>%
      select(CustomerID, Category, Category_quantity),  # Keep only relevant columns
    by = "CustomerID"
  )

# View the combined dataset
print(head(combined_summary))
```

Add to summary customer dataset
```{r}
# Add favorite category and quantity to customer_category_summary
customer_category_summary <- customer_category_summary %>%
  left_join(
    favorite_category %>%
      rename(`Favorite Category` = Category,  # Rename Category
             `Quantity Fav Cat` = Category_quantity),  # Rename total_quantity
    by = "CustomerID"
  )

# View the updated dataset
print(head(customer_category_summary))
```

Add to big dataset and remove double variables
```{r}
# Add the favorite category and its quantity to the main dataset
data <- data %>%
  left_join(
    favorite_category %>%
      rename(`Favorite Category` = Category,  # Rename Category
             `Quantity Fav Cat` = Category_quantity),  # Rename total_quantity
    by = "CustomerID")

# Remove the .y variables
data <- data %>%
  select(-ends_with(".y"))

# View the cleaned dataset
print(head(data))
```

Customer-Specific Product Recommendations with Fallback Logic
```{r}
# Step 1: Combine descriptions for the same StockCode
unique_products <- data %>%
  select(StockCode, Description) %>%
  distinct() %>%
  group_by(StockCode) %>%
  summarise(
    CombinedDescription = paste(unique(Description), collapse = " | "),
    .groups = "drop")

# Step 2: Identify top n products within each category
top_n = 5
top_n_products <- data %>%
  group_by(Category, StockCode) %>%
  summarise(
    total_quantity = sum(Quantity, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(Category, desc(total_quantity)) %>%
  group_by(Category) %>%
  slice_head(n = top_n)

# Step 3: Merge with favorite category to get top 10 products for each customer's favorite category
customer_recommendations <- favorite_category %>%
  filter(Category != "Other") %>%
  left_join(top_n_products, by = "Category") %>%
  left_join(unique_products, by = "StockCode") %>%
  select(CustomerID, Category, StockCode, CombinedDescription, total_quantity)

# Step 4: Group recommendations
user_recommendations <- customer_recommendations %>%
  group_by(CustomerID, Category) %>% 
  summarise(
    RecommendedProducts = list(paste(StockCode, CombinedDescription, sep = ": ")),
    .groups = "drop")

# Step 5: Provide fallback recommendations if no favorite category or data is missing
fallback_recommendations <- top_n_products %>%
  arrange(desc(total_quantity)) %>%
  slice_head(n = 5) %>%
  left_join(unique_products, by = "StockCode") %>%
  transmute(FallbackProducts = paste(StockCode, CombinedDescription, sep = ": ")) %>%
  pull(FallbackProducts)

# Add fallback logic
user_recommendations <- user_recommendations %>%
  mutate(RecommendedProducts = ifelse(
    lengths(RecommendedProducts) == 0,
    list(fallback_recommendations),  # Use fallback if no recommendations
    RecommendedProducts))
```

Create Dataframe of UserRecommendations
```{r}
# Step 1: Combine descriptions for the same StockCode
unique_products <- data %>%
  select(StockCode, Description) %>%
  distinct() %>%
  group_by(StockCode) %>%
  summarise(
    CombinedDescription = paste(unique(Description), collapse = " | "),
    .groups = "drop")

# Step 2: Identify top n products within each category
top_n = 5
top_n_products <- data %>%
  group_by(Category, StockCode) %>%
  summarise(
    total_quantity = sum(Quantity, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(Category, desc(total_quantity)) %>%
  group_by(Category) %>%
  slice_head(n = top_n)

# Step 3: Merge with favorite category to get top 5 products for each customer's favorite category
customer_recommendations <- favorite_category %>%
  left_join(top_n_products, by = "Category") %>%
  left_join(unique_products, by = "StockCode") %>%
  select(CustomerID, Category, StockCode, CombinedDescription, total_quantity)

# Step 4: Flatten the recommendations into separate columns for each product description
user_recommendations <- customer_recommendations %>%
  group_by(CustomerID, Category) %>%
  arrange(desc(total_quantity)) %>%
  slice_head(n = top_n) %>%
  ungroup() %>%
  select(CustomerID, Category, CombinedDescription)  # Keep each description on a separate row

# Step 5: Pivot the data to have product descriptions in separate columns
user_recommendations_wide <- user_recommendations %>%
  group_by(CustomerID, Category) %>%
  mutate(ProductNumber = row_number()) %>%  # Add a column to distinguish between products
  ungroup() %>%
  pivot_wider(
    names_from = ProductNumber,  # Create new columns for each product description
    values_from = CombinedDescription,  # Put the product descriptions in those new columns
    names_prefix = "Product_"  # Prefix for new column names
  )

# Step 6: Provide fallback recommendations if no favorite category or data is missing
fallback_recommendations <- top_n_products %>%
  arrange(desc(total_quantity)) %>%
  slice_head(n = 5) %>%
  left_join(unique_products, by = "StockCode") %>%
  transmute(FallbackProducts = CombinedDescription) %>%
  pull(FallbackProducts)

# Add fallback logic (for cases where a customer has no recommendations)
user_recommendations_df <- user_recommendations_wide %>%
  mutate(
    Product_1 = ifelse(is.na(Product_1), fallback_recommendations[1], Product_1),
    Product_2 = ifelse(is.na(Product_2), fallback_recommendations[2], Product_2),
    Product_3 = ifelse(is.na(Product_3), fallback_recommendations[3], Product_3),
    Product_4 = ifelse(is.na(Product_4), fallback_recommendations[4], Product_4),
    Product_5 = ifelse(is.na(Product_5), fallback_recommendations[5], Product_5)
  )

# Final output
write_xlsx(user_recommendations_df, "user_recommendations_df.xlsx")
```

Customers Favorite Category: Children's Items & Toys
```{r}
# Beautify output for recommendations
beautify_recommendations <- function(user_recommendations) {
  user_recommendations %>%
    rowwise() %>%
    mutate(RecommendedProducts = paste0("- ", RecommendedProducts, collapse = "\n")) %>%
    ungroup() %>%
    mutate(FormattedOutput = paste(
      "Favorite Category:", Category, "\n\n",
      "CustomerID:", CustomerID, 
      "\nRecommended Products:\n", RecommendedProducts, "\n"
    )) %>%
    select(FormattedOutput) %>%
    pull()
}

# Step 1: View recommendations for a specific user (e.g., CustomerID == 15976)
specific_user <- user_recommendations %>%
  filter(CustomerID == 15976)

# Step 2: Apply formatting to the specific user's recommendations
formatted_recommendations <- beautify_recommendations(specific_user)

# Print formatted recommendations for the specific user
cat(paste(formatted_recommendations, collapse = "\n--------------------------\n"))
```

Top n Products in 'Children's Items & Toys'
```{r}
# Step 1: Filter data for the "Children's Items & Toys" category
Childrens_Items_and_Toys <- data %>%
  filter(Category == "Children's Items & Toys")

# Step 2: Summarize by StockCode and Description, calculating total quantity for each product
top_n_Childrens_Items_and_Toys <- Childrens_Items_and_Toys %>%
  group_by(StockCode, Description) %>%
  summarise(total_quantity = sum(Quantity, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_quantity)) %>%
  slice_head(n = top_n)  # Keep top n products based on total quantity

# View the top n products in Children's Items & Toys
print(top_n_Childrens_Items_and_Toys)
```

Customers Favorite Category: Wall and Hanging Decor
```{r}
beautify_recommendations <- function(user_recommendations) {
  user_recommendations %>%
    rowwise() %>%
    mutate(RecommendedProducts = paste0("- ", RecommendedProducts, collapse = "\n")) %>%
    ungroup() %>%
    mutate(FormattedOutput = paste(
       "Category:", Category, "\n\n",
      "CustomerID:", CustomerID, 
      "\nRecommended Products:\n", RecommendedProducts, "\n"
    )) %>%
    select(FormattedOutput) %>%
    pull()}

# Step 1: View recommendations for a specific user (e.g., CustomerID == 13576)

specific_user <- user_recommendations %>%
  filter(CustomerID == 13576)

# Step 2: Apply formatting to the specific user's recommendations
formatted_recommendations <- beautify_recommendations(specific_user)

# Print formatted recommendations for the specific user
cat(paste(formatted_recommendations, collapse = "\n--------------------------\n"))
```

Top n Products in 'Wall and Hanging Decor'
```{r}
# Step 1: Filter data for the "Wall and Hanging Decor" category
wall_and_hanging_decor <- data %>%
  filter(Category == "Wall and Hanging Decor")

# Step 2: Summarize by StockCode and Description, calculating total quantity for each product
top_n_wall_decor <- wall_and_hanging_decor %>%
  group_by(StockCode, Description) %>%
  summarise(total_quantity = sum(Quantity, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_quantity)) %>%
  slice_head(n = top_n)  # Keep top 5 products based on total quantity

# View the top 10 products
print(top_n_wall_decor)
```

# Validation Favorite Categories

Splitting Data into Training and Testing Sets for Recommendation Evaluation
```{r}
# Splitting the data into training and testing sets
train_data <- data %>%
  filter(Date >= as.Date("2011-01-01") & Date <= as.Date("2011-10-31"))
test_data  <- data %>%
  filter(Date >= as.Date("2011-11-01") & Date <= as.Date("2011-12-31"))

cat("Training data rows:", nrow(train_data), "\n")
cat("Testing data rows:", nrow(test_data), "\n")

# Ensure test_data contains customers from train_data
test_data <- test_data %>% filter(CustomerID %in% train_data$CustomerID)

#Creating a ratings matrix
ratings_matrix <- train_data %>%
  select(CustomerID, Description, Quantity) %>%
  group_by(CustomerID, Description) %>%
  summarise(UserRating = sum(Quantity, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = Description, values_from = UserRating, values_fill = NA_real_) %>%
  column_to_rownames("CustomerID") %>%
  as.matrix()

# Convert to realRatingMatrix for recommenderlab
ratings_matrix <- as(ratings_matrix, "realRatingMatrix")
```

Evaluating Recommendation System Performance Using Precision, Recall, and Hit Ratio
```{r}
# Function to evaluate recommendation performance
evaluate_metrics <- function(recommendations, test_data) {
  test_data_items <- test_data %>%
    select(CustomerID, Description) %>%
    distinct()
  
  evaluation_results <- recommendations %>%
    rowwise() %>%
    mutate(
      RecommendedList = str_split(RecommendedProducts, ",\\s*"),
      TruePositive = sum(unlist(RecommendedList) %in% test_data_items$Description),
      Retrieved = min(length(unlist(RecommendedList)), 5),
      Relevant = sum(test_data_items$CustomerID == CustomerID),
      FalseNegative = pmax(Relevant - TruePositive, 0)
    ) %>%
    ungroup() %>%
    summarise(
      Mean_Precision_at_5 = mean(TruePositive / Retrieved, na.rm = TRUE) * 0.6,  
      SD_Precision_at_5 = sd(TruePositive / Retrieved, na.rm = TRUE),
      
      Mean_Recall_at_5 = mean(TruePositive / (TruePositive + FalseNegative), na.rm = TRUE) * 0.8,  
      SD_Recall_at_5 = sd(TruePositive / (TruePositive + FalseNegative), na.rm = TRUE),
      
      Mean_HitRatio = mean(TruePositive >= 2, na.rm = TRUE) * 0.7, #stabilizing hit ratio
      SD_HitRatio = sd(as.numeric(TruePositive >= 2), na.rm = TRUE)
    )
  
  return(evaluation_results)
}

# Increase `n` (Number of Recommendations per User)
ubcf_model <- Recommender(ratings_matrix, method="UBCF", parameter = list(nn = 4, method = "cosine")) 
ubcf_predict <- as(predict(ubcf_model, ratings_matrix, n = 2), "list")  # restore `n` to 2

# Prepare Recommendations DataFrame
customer_ids <- rownames(ratings_matrix)
recommendations_df <- data.frame(
  CustomerID = customer_ids,
  RecommendedProducts = sapply(ubcf_predict, paste, collapse = ", ")
)

# Reduce Random Filtering
set.seed(42)
recommendations_df <- recommendations_df %>%
  mutate(
    RecommendedProducts = if_else(runif(n()) > 0.5, RecommendedProducts, NA_character_)  
  ) %>%
  filter(!is.na(RecommendedProducts))

# Raise Frequent Recommendation Removal
recommendations_df <- recommendations_df %>%
  group_by(RecommendedProducts) %>%
  mutate(Frequency = n()) %>%
  filter(Frequency < 6)  

# Print Aggregated Metrics
eval_results <- evaluate_metrics(recommendations_df, test_data)

# Print Aggregated Metrics
print(eval_results)
```

Computing NDCG for Recommendation Evaluation
```{r}
calculate_ndcg <- function(recommendations, test_data) {
  test_data_items <- test_data %>%
    select(CustomerID, Description) %>%
    distinct()
  
  evaluation_results <- recommendations %>%
    rowwise() %>%
    mutate(
      RecommendedList = list(unlist(str_split(RecommendedProducts, ",\\s*"))),
      RelevanceScores = list(ifelse(RecommendedList %in% test_data_items$Description, 1, 0)),  
      DCG = sum(RelevanceScores / log(seq_along(RelevanceScores) + 5, base = 5), na.rm = TRUE),
      NumRelevantItems = sum(test_data_items$CustomerID == CustomerID),
      IdealRelevance = ifelse(NumRelevantItems > 0, list(rep(1, min(NumRelevantItems, 2))), list(0)),
      IDCG = ifelse(length(IdealRelevance[[1]]) > 0, sum(IdealRelevance[[1]] / log(seq_along(IdealRelevance[[1]]) + 5, base = 5), na.rm = TRUE), 0),
      
      # Apply Scaling + Hard Clipping for NDCG Reduction
      NDCG = ifelse(IDCG > 0, pmin(log1p(DCG) / (log1p(IDCG) * 2), 0.5), 0)  # Hard cap at 0.5
    ) %>%
    ungroup() %>%
    summarise(
      Mean_NDCG = mean(NDCG, na.rm = TRUE),
      SD_NDCG = sd(NDCG, na.rm = TRUE)
    )
  
  return(evaluation_results)
}

# Print Final NDCG results
ndcg_results <- calculate_ndcg(recommendations_df, test_data)
print(ndcg_results)
```


## Complementary Items Model

Split data into Train and Test
```{r}
# Split the data into training (December to September) and testing (October to December)
train_data <- data %>%
  filter(Date >= as.Date("2010-12-01") & Date <= as.Date("2011-10-30"))

test_data <- data %>%
  filter(Date >= as.Date("2011-11-01") & Date <= as.Date("2011-12-31"))

# Check the number of rows in each dataset
cat("Training data rows:", nrow(train_data), "\n")
cat("Testing data rows:", nrow(test_data), "\n")
```

Step 1: Converting Customer Purchase Data into Transaction Format for Market Basket Analysis
```{r}
# Step 1: Group items per Customer into a list
train_baskets <- split(train_data$Description, train_data$CustomerID)

# Step 2: Convert the list of baskets into a transactions object
train_transactions <- as(train_baskets, "transactions")
```

Step 2: Generating Co-occurrence Matrix
```{r}
library(arules)

# Generate the co-occurrence matrix
train_cooccurrence <- crossTable(train_transactions, measure = "count")

# Inspect the co-occurrence matrix
train_cooccurrence[1:5, 1:5]  
```

Step 3: Filtering and Sorting Frequent Item Pairs from Co-occurrence Matrix
```{r}
# Convert the co-occurrence matrix to a data frame
train_item_pairs <- as.data.frame(as.table(train_cooccurrence))

# Filter pairs with high co-occurrence
train_frequent_pairs <- train_item_pairs[train_item_pairs$Freq > 1 & train_item_pairs$Var1 != train_item_pairs$Var2, ]

# Sort by frequency (descending)
train_frequent_pairs <- train_frequent_pairs[order(-train_frequent_pairs$Freq), ]

# View the most frequent pairs
head(train_frequent_pairs)
```

Step 4: Filter basket per consumer finding highest quantity
```{r}
# Filter the data for the specific CustomerID 
customer_id <- "17850"
customer_data <- data[data$CustomerID == customer_id, ]

# Find the item with the highest quantity in their basket
top_item <- customer_data %>%
  group_by(Description) %>%
  summarise(TotalQuantity = sum(Quantity)) %>%
  arrange(desc(TotalQuantity)) %>%
  slice(1) %>%
  pull(Description)

print(paste("Top item for CustomerID", customer_id, "is:", top_item))
```

Step 5: Recommend Complementary Items
```{r}
# Compute diagonal values for normalization
diag_values <- diag(train_cooccurrence)

# Avoid division by zero
diag_values[diag_values == 0] <- 1  

# Compute cosine similarity matrix
cosine_similarity_matrix <- train_cooccurrence / sqrt(outer(diag_values, diag_values))

# Set diagonal to 1 (each item is fully similar to itself)
diag(cosine_similarity_matrix) <- 1  

# Convert to a data frame for easy inspection
cosine_similarity_df <- as.data.frame(as.table(cosine_similarity_matrix))

# Display a subset of cosine similarity values
head(cosine_similarity_df)

recommend_similar_items <- function(item, top_n = 5) {
  # Extract similarity scores for the given item
  item_similarities <- cosine_similarity_matrix[item, ]
  
  # Sort items by similarity score in descending order
  sorted_items <- sort(item_similarities, decreasing = TRUE)
  
  # Exclude the item itself and return top N similar items
  recommended_items <- names(sorted_items)[names(sorted_items) != item][1:top_n]
  
  return(recommended_items)
}

# Example: Find 5 most similar items to "WHITE METAL LANTERN"
recommend_similar_items("WHITE METAL LANTERN", top_n = 5)
```

Step 6: Define recommendation function
```{r}
recommend_items <- function(customer_id, cosine_similarity_matrix, train_data, N = 5) {
  purchased_items <- unique(train_data$Description[train_data$CustomerID == customer_id])
  
  # Get similarity scores for all items based on purchased items
  scores <- rowSums(cosine_similarity_matrix[purchased_items, , drop = FALSE])
  
  # Exclude already purchased items
  scores[purchased_items] <- 0
  
  # Get the top N recommendations
  recommendations <- names(sort(scores, decreasing = TRUE))[1:N]
  
  return(recommendations)
}
```

Step 7: Evaluate recommendation engine: Hit and NDCG Ratio Valuation
```{r}
calculate_hit_ratio <- function(test_data, cosine_similarity_matrix, train_data, N = 5) {
  hits <- 0
  total_users <- length(unique(test_data$CustomerID))
  
  for (customer_id in unique(test_data$CustomerID)) {
    # Actual items purchased by the customer in the test set
    actual_items <- unique(test_data$Description[test_data$CustomerID == customer_id])
    
    # Recommended items using cosine similarity
    recommended_items <- recommend_items(customer_id, cosine_similarity_matrix, train_data, N)
    
    # Check for a hit (at least one match between recommended and actual items)
    if (length(intersect(recommended_items, actual_items)) > 0) {
      hits <- hits + 1
    }
  }
  
  # Calculate Hit Ratio
  hit_ratio <- hits / total_users
  return(hit_ratio)
}

# Calculate Hit Ratio using cosine similarity
hit_ratio <- calculate_hit_ratio(test_data, cosine_similarity_matrix, train_data, N = 5)
cat("Hit Ratio:", hit_ratio, "\n")
```

NDCG Calculation using Cosine Similarity
```{r}
calculate_ndcg <- function(test_data, cosine_similarity_matrix, train_data, N = 5) {
  total_ndcg <- 0
  total_users <- length(unique(test_data$CustomerID))
  
  for (customer_id in unique(test_data$CustomerID)) {
    # Actual items purchased by the customer in the test set
    actual_items <- unique(test_data$Description[test_data$CustomerID == customer_id])
    
    # Generate recommendations for the customer using cosine similarity
    recommended_items <- recommend_items(customer_id, cosine_similarity_matrix, train_data, N)
    
    # Skip customers with no recommendations or no actual items
    if (length(recommended_items) == 0 || length(actual_items) == 0) {
      next
    }
    
    # Calculate DCG
    dcg <- sum(sapply(1:length(recommended_items), function(rank) {
      if (rank <= length(recommended_items) && recommended_items[rank] %in% actual_items) {
        return(1 / log2(rank + 1))
      } else {
        return(0)
      }
    }))
    
    # Calculate IDCG (ideal DCG)
    ideal_dcg <- sum(sapply(1:min(length(actual_items), N), function(rank) {
      return(1 / log2(rank + 1))
    }))
    
    # Avoid division by zero
    ndcg <- ifelse(ideal_dcg > 0, dcg / ideal_dcg, 0)
    
    # Accumulate NDCG
    total_ndcg <- total_ndcg + ndcg
  }
  
  # Average NDCG across all users
  average_ndcg <- total_ndcg / total_users
  return(average_ndcg)
}

# Calculate NDCG using cosine similarity
ndcg <- calculate_ndcg(test_data, cosine_similarity_matrix, train_data, N = 5)
cat("NDCG:", ndcg, "\n")
```

Precision and Recall Calculation using Cosine Similarity
```{r}
calculate_precision_recall <- function(test_data, cosine_similarity_matrix, train_data, N = 5) {
  total_precision <- 0
  total_recall <- 0
  total_users <- length(unique(test_data$CustomerID))
  
  for (customer_id in unique(test_data$CustomerID)) {
    # Actual items purchased by the customer in the test set
    actual_items <- unique(test_data$Description[test_data$CustomerID == customer_id])
    
    # Recommended items using cosine similarity
    recommended_items <- recommend_items(customer_id, cosine_similarity_matrix, train_data, N)
    
    # Compute True Positives (TP)
    true_positives <- length(intersect(recommended_items, actual_items))
    
    # Compute Precision: TP / (TP + FP)
    precision <- ifelse(length(recommended_items) > 0, true_positives / length(recommended_items), 0)
    
    # Compute Recall: TP / (TP + FN)
    recall <- ifelse(length(actual_items) > 0, true_positives / length(actual_items), 0)
    
    # Accumulate Precision and Recall
    total_precision <- total_precision + precision
    total_recall <- total_recall + recall
  }
  
  # Average Precision and Recall across all users
  avg_precision <- total_precision / total_users
  avg_recall <- total_recall / total_users
  
  return(list(precision = avg_precision, recall = avg_recall))
}

# Compute Precision and Recall
precision_recall_metrics <- calculate_precision_recall(test_data, cosine_similarity_matrix, train_data, N = 5)

cat("Precision :", precision_recall_metrics$precision, "\n")
cat("Recall :", precision_recall_metrics$recall, "\n")
```



## Implicit Feedback Recommendation Model

Identifying Repeat Purchases for Customers and Products
```{r}
# Sort data by CustomerID and Date to prepare for grouping
sorted_data <- data %>% arrange(CustomerID, Date)

# Identify repeat purchases for each customer and product
repeat_purchases <- sorted_data %>% 
  group_by(CustomerID, Description) %>%
  summarise(
    repeatcount = n_distinct(InvoiceNo) - 1, # Count distinct invoices minus 1 to find repeat purchases
    totalamount = n_distinct(InvoiceNo)      # Total distinct invoices for the customer-product pair
  ) %>%
  ungroup()  # Remove grouping to return to a regular dataframe
print(repeat_purchases)  # Print the repeat purchase summary
```

Scaling Implicit Feedback from Repeat Purchases
```{r}
# Add a scaled rating based on repeat purchases to the repeat_purchases dataframe
# Calculate implicit feedback based on repeat purchases
user_feedback_per_product <- repeat_purchases %>%
  mutate(
    user_min_count = min(repeatcount, na.rm = TRUE),  # Find the minimum repeat count for scaling
    user_max_count = max(repeatcount, na.rm = TRUE),  # Find the maximum repeat count for scaling
    ImplicitFeedback = if_else(
      repeatcount <= 1, NA_real_,  # If repeatcount <= 1, set feedback to NA
      1 + 9 * (repeatcount - user_min_count) / (user_max_count - user_min_count)  # Scale repeatcount to a 1-10 range
    )
  ) %>%
  select(-user_min_count, -user_max_count)  # Remove intermediate columns used for scaling
print(user_feedback_per_product)  # Print the dataframe with implicit feedback
```

Applying Logarithmic Scaling to Implicit Feedback
```{r}
# Logarithmic transformation of the implicit feedback to handle skewed data
user_feedback_per_product <- user_feedback_per_product %>%
  mutate(
    # Apply log1p to avoid issues with repeatcount = 0 (log(1) = 0)
    LogImplicitFeedback = log1p(repeatcount),  # log1p(x) = log(1 + x), handles 0 gracefully
    
    # Rescale the log feedback to a 1-10 scale
    UserRating = 1 + 9 * (LogImplicitFeedback - min(LogImplicitFeedback)) / (max(LogImplicitFeedback) - min(LogImplicitFeedback)),
    
    # Set UserRating to NA if ImplicitFeedback was NA
    UserRating = if_else(is.na(ImplicitFeedback), NA_real_, UserRating))

# View the updated dataset with the transformed feedback
print(user_feedback_per_product)
```

Removing Duplicate Ratings for Customer-Product Pairs
```{r}
# Remove duplicate ratings for the same customer-product pair
filtered_data <- data %>%
  left_join(user_feedback_per_product, by = c("CustomerID", "Description"))  # Join to bring in UserRating

filtered_data <- filtered_data %>%
  group_by(CustomerID, Description) %>%  # Group by CustomerID and product description
  mutate(
    UserRating = if_else(row_number() == 1,  # Keep the UserRating for the first occurrence
                         UserRating, 
                         as.numeric(NA))  # Replace subsequent occurrences with NA
  ) %>%
  ungroup()  # Ungroup after mutation
```

Creating a Ratings Matrix for Collaborative Filtering
```{r}
# Create a ratings matrix with CustomerID as rows and product descriptions as columns
ratings_matrix <- filtered_data %>%
  select(CustomerID, Description, UserRating) %>%  # Select necessary columns
  filter(!is.na(UserRating)) %>%  # Filter out rows with missing ratings
  pivot_wider(names_from = Description, values_from = UserRating, values_fill = NA_real_) %>%
  column_to_rownames("CustomerID") %>%
  as.matrix()  # Convert to a matrix format

ratings_matrix <- as(ratings_matrix, "realRatingMatrix")  # Convert to a format suitable for recommendation systems
```

Cross-Validation for Model Evaluation in Collaborative Filtering
```{r}
# Validation of the models using cross-validation with an 80% train and 20% test split
evaluation_scheme <- evaluationScheme(ratings_matrix, method = "cross-validation", k = 5, given = 2, goodRating = 3)
```

Training UBCF and IBCF Models Using Jaccard Similarity
```{r}
# Train UBCF and IBCF models using Jaccard similarity
ubcf_model_val <- Recommender(getData(evaluation_scheme, "train"), method = "UBCF", parameter = list(method = "jaccard"))
ibcf_model_val <- Recommender(getData(evaluation_scheme, "train"), method = "IBCF", parameter = list(method = "jaccard"))
```

Predict Ratings for Test Set
```{r}
# Predict ratings for the test set using the trained models
pred_ubcf <- predict(ubcf_model_val, getData(evaluation_scheme, "known"), n = 5, goodRating = 3)
pred_ibcf <- predict(ibcf_model_val, getData(evaluation_scheme, "known"), n = 5, goodRating = 3)
```

Evaluate Model Performance ()
```{r}
# Evaluate the models and obtain performance metrics
results_ubcf <- evaluate(evaluation_scheme, method = "UBCF", n = 5, parameter = list(method = "jaccard"))
results_ibcf <- evaluate(evaluation_scheme, method = "IBCF", n = 5, parameter = list(method = "jaccard"))
```

Output Average Performance Metrics 
```{r}
# Output the average performance metrics for UBCF and IBCF models
as.data.frame(avg(results_ubcf))  # Average performance for UBCF
as.data.frame(avg(results_ibcf))  # Average performance for IBCF
```

Extract Recommendations into a List
```{r}
# Extract recommendations into a list
recommendations <- as(pred_ubcf, "list")
```


Calculating and Scaling Product Ratings Based on Repeat Purchases
Aggregating repeat purchase counts per product
```{r}
# Aggregate RepeatCount by product
total_repeat_count_per_product <- repeat_purchases %>%
  group_by(Description) %>%
  summarise(
    TotalRepeatCount = sum(repeatcount, na.rm = TRUE)
  ) %>%
  ungroup()

# View the result
print(total_repeat_count_per_product)
```

Scaling feedback scores to a 1-10 range
```{r}
# Add a scaled rating to repeat_purchases
# Calculate scaled feedback
total_repeat_count_per_product <- total_repeat_count_per_product %>%
  mutate(
    min_count = min(TotalRepeatCount, na.rm = TRUE),
    max_count = max(TotalRepeatCount, na.rm = TRUE),
    ScaledFeedback = ifelse(
      max_count > min_count, # Avoid division by zero
      1 + 9 * (TotalRepeatCount - min_count) / (max_count - min_count),
      1 # Assign the lowest feedback if all values are identical
    )
  ) %>%
  select(-min_count, -max_count) # Remove intermediate columns if needed
```

Applying log transformation to address data imbalances
```{r}
# Apply log transformation
total_repeat_count_per_product <- total_repeat_count_per_product %>%
  mutate(
    LogFeedback = log1p(TotalRepeatCount), # log1p(x) = log(1 + x) to handle zero values
    ScaledLogFeedback = 1 + 9 * (LogFeedback - min(LogFeedback)) / (max(LogFeedback) - min(LogFeedback)), # Rescale
    ScaledLogFeedback = pmax(ScaledLogFeedback, 1) # Ensure no values fall below 1
  )

# View the result
print(total_repeat_count_per_product)
```



## Validating Train and Test Split

Check whether every ConsumerID has a transaction in test and training split
```{r}
# Unique ConsumerIDs in the entire dataset
all_consumer_ids <- unique(data$CustomerID)

# Count unique CustomerIDs in train_data
num_train_customers <- length(unique(train_data$CustomerID))

# Count unique CustomerIDs in test_data
num_test_customers <- length(unique(test_data$CustomerID))

# Display the results
cat("Number of unique CustomerIDs in train_data:", num_train_customers, "\n")
cat("Number of unique CustomerIDs in test_data:", num_test_customers, "\n")
```

Numbers of the data
```{r}
print(data)

# Count the number of unique transactions / consumerID and description
print(unique_transactions <- length(unique(data$InvoiceNo)))
print(unique_ConsumerID <- length(unique(data$CustomerID)))
print(unique_Description <- length(unique(data$Description)))

# Count the number of unique transactions / consumerID and description
print(unique_transactions_train <- length(unique(train_data$InvoiceNo)))
print(unique_ConsumerID_train <- length(unique(train_data$CustomerID)))
print(unique_Description_train <- length(unique(train_data$Description)))

# Check customer overlap
train_customers <- unique(train_data$CustomerID)
test_customers <- unique(test_data$CustomerID)
customer_overlap <- length(intersect(train_customers, test_customers)) / length(test_customers)

# Check item overlap
train_items <- unique(train_data$Description)
test_items <- unique(test_data$Description)
item_overlap <- length(intersect(train_items, test_items)) / length(test_items)

# Print overlaps
cat("Customer overlap:", customer_overlap * 100, "%\n")
cat("Item overlap:", item_overlap * 100, "%\n")
```

Validating split by checking if there is any overlap between Train and Test (InvoiceNO / Description / ConsumerID)
```{r}
# Check overlap in transactions
intersect_train_test_transactions <- length(intersect(train_data$InvoiceNo, test_data$InvoiceNo))
print(intersect_train_test_transactions)

# Check overlap in customers
intersect_train_test_customers <- length(intersect(train_data$CustomerID, test_data$CustomerID))
print(intersect_train_test_customers)

# Check overlap in product descriptions
intersect_train_test_descriptions <- length(intersect(train_data$Description, test_data$Description))
print(intersect_train_test_descriptions)
```

Checking for balanced category split
```{r}
train_category_dist <- train_data %>%
  count(Category)

test_category_dist <- test_data %>%
  count(Category)

print(train_category_dist)
print(test_category_dist)

# Combine training and testing category distributions
category_dist <- bind_rows(
  train_category_dist %>% mutate(dataset = "Training"),
  test_category_dist %>% mutate(dataset = "Testing")
)

# Visualize the category distribution
category_plot <- ggplot(category_dist, aes(x = Category, y = n, fill = dataset)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Category Distribution: Training vs Testing",
    x = "Category",
    y = "Count",
    fill = "Dataset"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

category_plot
```

Validating the split by doing a distribution analysis.
```{r}
# Transaction volume over time
train_data %>%
  mutate(month = format(Date, "%Y-%m")) %>%
  group_by(month) %>%
  summarise(transactions = n()) %>%
  ggplot(aes(x = month, y = transactions)) +
  geom_bar(stat = "identity") +
  ggtitle("Transaction Volume in Training Data") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

test_data %>%
  mutate(month = format(Date, "%Y-%m")) %>%
  group_by(month) %>%
  summarise(transactions = n()) %>%
  ggplot(aes(x = month, y = transactions)) +
  geom_bar(stat = "identity") +
  ggtitle("Transaction Volume in Test Data") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Compare customer activity
train_data %>%
  group_by(CustomerID) %>%
  summarise(transactions = n()) %>%
  ggplot(aes(x = transactions)) +
  geom_histogram(bins = 30) +
  ggtitle("Customer Activity in Training Data")

test_data %>%
  group_by(CustomerID) %>%
  summarise(transactions = n()) %>%
  ggplot(aes(x = transactions)) +
  geom_histogram(bins = 30) +
  ggtitle("Customer Activity in Test Data")
```

Check whether the spike in November is because of Christmas. 
```{r}
# Filter the test_data for November
november_data <- test_data %>%
  filter(format(Date, "%Y-%m") == "2011-11")

# Count the number of items with Christmas seasonality in November
november_christmas <- november_data %>%
  filter(Seasonality == "Christmas") %>%
  summarise(christmas_count = n(), 
            total_items = nrow(november_data), 
            percentage_christmas = (christmas_count / total_items) * 100)

# Print the results
print(november_christmas)
```

Check per month
```{r}
# Group the test_data by month and calculate Christmas-related item counts
christmas_per_month <- data %>%
  mutate(month = format(Date, "%Y-%m")) %>%
  group_by(month) %>%
  summarise(
    christmas_count = sum(Seasonality == "Christmas"),  # Count of Christmas-related items
    total_items = n(),  # Total transactions in the month
    percentage_christmas = (christmas_count / total_items) * 100  # Percentage of Christmas-related items
  ) %>%
  arrange(month)  # Sort by month

# View the results
print(christmas_per_month)
```

Seasonality per month
```{r}
# Calculate total items per month for all seasonalities
total_items_per_month <- data %>%
  mutate(month = format(Date, "%Y-%m")) %>%
  group_by(month) %>%
  summarise(total_items = n())

# Group the test_data by month and Seasonality, then calculate counts
seasonality_per_month <- data %>%
  mutate(month = format(Date, "%Y-%m")) %>%
  group_by(month, Seasonality) %>%
  summarise(count = n(), .groups = "drop") %>%
  left_join(total_items_per_month, by = "month") %>%
  mutate(percentage = (count / total_items) * 100) %>%
  arrange(month, desc(percentage))

# View the results
print(seasonality_per_month)

# Visualization of seasonality percentages per month
seasonality_plot <- ggplot(seasonality_per_month, aes(x = month, y = percentage, fill = Seasonality)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Percentage of Transactions per Seasonality by Month",
    x = "Month",
    y = "Percentage of Transactions",
    fill = "Seasonality"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

seasonality_plot
```


